\huge{Sexual selection improves population fitness: a systematic review and meta-analysis}

### 2019-01-03 Response_to_Nature_Comms.Rmd

---
title: 'Sexual selection improves population fitness: a systematic review and meta-analysis'
output:
  pdf_document: default
  word_document: default
subtitle: Response to Decision Letter
bibliography: Manuscripts/Bibliography/Bibliography.bib
---

@@

We agree that alone, these 6 studies do not provide strong evidence the mean is positive, although we note that some of the precisely measured effect sizes, and one of the imprecise ones, are positive, so this borderline significant result is not too anomalous.

Secondly, the effect size predictions were made using a single model that incorporated hundreds of other effect sizes, which allows the model is allowed to 'share information' between the different fitness measures. For example, if the overall mean effect size is positive, then smaller subsets of the dataset (like the attractiveness subset, with n = 6 studies) will tend to be made slightly more positive. That is, in the absence of precise information about the true average effect size for studies of attractiveness, the model uses some information provided by the global average.

>*Thirdly, type of trait measured explained 35% variance, but the authors do not explore this any further. However, examination of Fig. 1 suggest that some indirect/ambiguous fitness measures account for much of this heterogeneity and they generally have higher average effect sizes that direct measures (except for immunity). I’d like to see if the authors recover their main result if they only direct measures.*

In the manuscript, we culled the dataset to only include the "direct" and "indirect" measures of popualtion fitness (this decision was taken to maximise the use of the data). If we were to eliminate indirect measures, it would leave us with n = **XX** "direct" effect sizes in our main analysis. To facilitate easy comparison, we present the new and old results below: one can see that two things have changed (in terms of statistical significance). Firstly, for the "direct only" model the predicted effect size for males in stressful environments is now significantly negative (*Hedges' g* = -0.45; 95 % CIs = -0.70 to -0.19). However, because in the revised dataset we have reclassified male reproductive success as an ambiguous measure of fitness, there are only two effect sizes for males in stressful environments; making any conclusions drawn from this finding problematic. Secondly, the predicted effect size for females in benign environments in the "direct only model" are non-significantly positive, unlike the "direct and indirect" model, however, in both cases the positive effect of sexual selection on fitness is magnified for females in stressful environments. To avoid reader fatigue in our manuscript (since we already chop and change the dataset in many different ways), we have retained the analysis to include both the "indirect and direct" measures. The graph on the right, and the table below show the main results recovered from only "direct" measurements for reference.

**Luke: Might need to argue use of both indirect and direct more. Basically the results **

@@

# lternatively you can obtain posterior samples manually.
post.direct <- (posterior_samples(brms.direct, 
                           pars = c("b_Intercept", "b_SexB", "b_SexF", 
                                    "b_EnvironmentStressed", "b_SexB:EnvironmentStressed", 
                                    "b_SexF:EnvironmentStressed")) %>%
           mutate(both_benign = b_Intercept + b_SexB,
                  both_stressful = b_Intercept + b_SexB + b_EnvironmentStressed + `b_SexB:EnvironmentStressed`,
                  male_benign = b_Intercept,
                  male_stressful = b_Intercept + b_EnvironmentStressed,
                  female_benign = b_Intercept + b_SexF,
                  female_stressful = b_Intercept + b_SexF + b_EnvironmentStressed + `b_SexF:EnvironmentStressed`))[,-(1:6)]

# Add columns for Environment and Sex
post.direct <- as.data.frame(t(post.direct))
post.direct$Sex <- c("Both", "Both", "Male", "Male", "Female", "Female")
post.direct$Environment <- c("Benign", "Stressful", "Benign", "Stressful", "Benign", "Stressful")

#Clean up dataframe
post.direct <- melt(post.direct, id = c("Sex", "Environment"))
post.direct$variable <- NULL

#Plot the posterior values from the Bayesian model as density ridges
pd <- position_dodgev(height = 0.3)
posterior.direct.plot <- post.direct %>% mutate(Sex = factor(Sex, levels = c("Male", "Both", "Female"))) %>% ggplot()+
  stat_density_ridges(aes(x=value, y = Environment, fill = Sex), alpha = 0.65, scale = 0.6, position = position_nudge(y = 0.15), height = 10, show.legend = F, quantile_lines = T, quantiles = 2)+
  geom_vline(xintercept = 0, linetype = 2, colour = "black") + 
  ylab("Environment\n")+
  xlab("\nEffect Size (Hedges' g)")+
  # scale_fill_manual(values = c("Male" = "#e41a1c", "Female" = "#377eb8", "Both" = "#4daf4a"))+
  # scale_color_manual(values = c("Male" = "#e41a1c", "Female" = "#377eb8", "Both" = "#4daf4a"))+
  scale_fill_manual(values = c("Male" = "#ff7f00", "Female" = "#984ea3", "Both" = "#4daf4a"))+
  scale_color_manual(values = c("Male" = "#ff7f00", "Female" = "#984ea3", "Both" = "#4daf4a"))+
  scale_x_continuous(limits = c(-1.25, 1.25), breaks = c(-1, -.5, 0, 0.5, 1))+

@@

  geom_point(data = predictions.direct %>% mutate(Sex = factor(Sex, levels = c("Male", "Both", "Female"))), 
             aes(x = Prediction, y = Environment, size=n, fill = Sex), 
             shape=21, color = "grey20", position = pd) +

    guides(fill = guide_legend(reverse=T, override.aes = list(size = 7.5)))+
    scale_size(guide = 'none')+

  scale_y_discrete(expand=c(0.075, 0))

grid.arrange(both.plots + 
               guides(fill = FALSE)+
               ggtitle("Indirect And Direct\n")+ theme(plot.title = element_text(size = 18, face = "bold", colour = "Red")),
             both.direct.plots + ggtitle("Direct Only\n")+
               theme(axis.title.y = element_blank(), 
                     plot.title = element_text(size = 18, face = "bold", colour = "Red")), 
             nrow = 1, 
             widths = c(3,3.5))
```

@@

>*As a response, they come up with log response ratio (lnRR) - see*

>Hedges, L. V., J. Gurevitch, and P. S. Curtis. 1999. The meta-analysis of response ratios in experimental ecology. \textit{Ecology} 80:1150-1156.

>*I recommend that the authors use lnRR for their effect size for mean comparison as well as Hedge's $g$ to see the robustness of their conclusions for the mean.*

After reading those references, we agree that Hedges' *g* is imperfect, and so we checked the robustness of our meta-analysis using two alternative measures of effect size. Firstly, we present the log response ratio (*lnRR*) as suggested by this reviewer. Secondly, we present a modified version of Hedges' *g* that attempts to account for this heterogeneity issue [@bonett2009meta], named SMDH (standardized mean difference with heteroscedastic population variances in the two groups). There are several key findings from our new analyses using lnRR and SMDH:

@@

Very useful, thank you!

________________________________________________________

## Comments from Reviewer 3

>*This manuscript addresses the question of whether, on average, sexual selection has a net beneficial or detrimental effect on population fitness, using a meta-analysis of experimental evolution studies comparing the fitness of populations under different intensities of sexual selection. The analyses consider the effects of sexual selection on both the mean and variance of population fitness, and test whether these effects differ for stressful vs. benign environments, or depend on the measure of population fitness.*

>*These are issues of longstanding interest, and as there is now quite a wealth of experimental evolution studies addressing these questions a meta-analysis synthesising their findings is timely.*

@@

Please see our earlier response to Prof. Radwan: we do not consider male mating success (and related traits) to be correlates of population fitness, which is why the results differ between our strict analysis and the one that includes all the traits. Additionally, we chose to spend more time writing about the "direct" effects because our main focus is on population fitness. We agree that our secondary results are interesting as well, but we wanted to keep the manuscript focused and concise. Also, there is no theoretical controversy over how sexual selection should affect traits like male mating success; by contrast, the relationship between sexual selection and population fitness is contentious.

>*I have only a few more specific comments:*

>*Does your approach consider variation in the intensity of sexual selection exclusively on males? This is worth noting (e.g. statements such as that on line 98-99 might be amended to “Sexual selection on males significantly improved female fitness.”)*

Yes, we now also stress that we consider sexual selection on males, as suggested (changes are made throughout the manuscript).

@@

>*Experiments manipulating the sex ratio to alter sexual selection might simultaneously decrease sexual selection on males while increasing sexual selection on females. Would this be classified as reduced sexual selection in your analyses? I don’t think you included this aspect of study design as a moderator variable in any of your analyses – given that close to half of the effects you include come from “alternative manipulations” (line 70) perhaps you could test whether this affects effect size?*

Some of the studies used the enforced monogamy design (as in Holland and Rice 2002), where one compares monogamous family groups with polyandrous ones (i.e. multiple males and one female), and so these manipulate sexual selection on males only, and eliminate sexual selection on females. Others instead manipulated the adult sex ratio (e.g. 3 males 1 female, 2 of each, or 3 females 1 male); these studies manipulate sexual selection on both sexes simultaneously. For the latter type, we considered the strength of sexual selection on males to be variable of interest, though it is confounded with the strength of sexual seleciton on females. To test if there was any difference in the effect sizes induced by these two types of manipulation, we now include 'manipulation type' (named "Enforced.Monogamy" in the Supplementary Material and R code) as a moderator variable: there was no effect in our preliminary model (**see above comment to Reviewer 1 and the revised Table S10**).

@@

>*It seems possible that sexual selection on females, and not only on males, could affect population fitness – potentially even more directly than sexual selection on males. And might differing extents of sexual selection in each sex interact with the differences you saw in fitness measured in males vs. females?*

It is possible that sexual selection on females on affects population fitness, although to our knowledge no-one has considered or modelled this before. Also, we are not a aware of any experiments that have manipulated the opportunity for sexual selection on females independently of any other factors (e.g. sexual selection on males; see above comment). So, we have opted not to discuss or analyse this further, as we think the data do not yet allow for a robust test.

>*L111-118 It seems a little odd that there is no residual heterogeneity in your $I^2$ estimates. From the supplemental information it is not entirely clear to me how you have adapted the function that is under development in “metaAidR”, but can you double check that you have appropriately incorporated the residual variance into these estimates?*

The results are odd, however to the best of our knowledge they are accurate. The REML model estimates the variance component ($\sigma^2$) of the "taxon" random effect as very low (3.38e-09), compared to the other random effects ("Study ID"" = 1.93e-01 and "fitness component" = 1.28e-01). A low value for $\sigma^2$ is the reason $I^2$ is estimated as zero. A potential reason for low heterogeneity is that there are only six taxon levels (campared to 53 studies and 13 fitness components).

**Check this please Justin - I could not see what you changed in the R script**
**Luke: I think the function was still in dev when i forked it and there was a bug that stopped in allowing me to get I2 for all three random effects (only last two), I changed the code so it recognised all three random effects. Not sure if the answer I provided makes sense**

@@

>*L307-311 Please clarify here that lnCVR was calculated as the ratio $ln(CVfitness_{SS}/CVfitness_{no SS})$*

We now clarify this as suggested.

>*Fig. 3 Please explain the dashed red and black lines in the figure caption.*

We now clarify the confidence intervals on the funnel plot as suggested.

@@

>*Tables 1, 2 Please check – the parameter estimates for ‘Female sex’ and test for ‘Female > Male’ are bolded inconsistently between the REML and Bayesian models where these estimates are identical.*

The revised manuscript no longer has this disparity between REML and Bayesian estimates. However we noticed that Bayesian estimates had slightly wider confidence intervals than REML models leading to differences in statistical significance (e.g. Table S6).

### 2019-01-03 Manuscripts/Manuscript_MetaAnalysis.Rmd
updated supp info

---
title: '\huge{Sexual selection improves population fitness: a meta-analysis}'
author: 
- Justin G. Cally\textsuperscript{§*}, Devi Stuart-Fox\textsuperscript{§} and Luke Holman\textsuperscript{§}
- 
- '\textsuperscript{*}\textit{justin.g.cally@gmail.com}'
output:
  pdf_document: default
  html_document: default
  word_document:
      pandoc_args: ["--smart"]
bibliography: Bibliography/Bibliography.bib
subparagraph: yes
csl: Bibliography/NComms.csl
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[LO,LE]{\textsl{\leftmark}}
  \rhead[]{Sexual selection improves fitness} \usepackage{titlefoot}
  \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno} \linenumbers \usepackage{setspace}\onehalfspacing
    \usepackage{titlesec} \titleformat{\section}{\normalfont\scshape\LARGE}{\thesection}{1em}{} \titleformat{\subsection}{\normalfont\Large}{\thesubsection}{1em}{} \titleformat{\subsubsection}{\normalfont\large}{\thesubsubsection}{1em}{} \usepackage{abstract} \renewcommand{\abstractnamefont}{\normalfont\Large} 
    \usepackage{hyperref}\hypersetup{colorlinks=false, linkcolor=blue, filecolor=blue, urlcolor=blue} 
    \usepackage{color,soul}
    \raggedbottom    
abstract: Sexual selection has manifold ecological and evolutionary consequences, making its net effect on population fitness difficult to predict from first principles. Instead, one can empirically determine whether sexual selection increases or decreases population fitness by  manipulating the opportunity for sexual selection in replicate populations, allow evolution to proceed, and then compare the evolved average fitness between the treatment groups. Here, we synthesise 459 effect sizes from 65 pertinent experimental evolution studies using meta-analysis. We find that sexual selection tends to elevate the mean and reduce the variance for many fitness traits, especially in females, and for populations evolving under stressful conditions. We conclude that the beneficial population-level consequences of sexual selection typically outweigh the harmful ones and that the effects of sexual selection on total selection, genetic variance, and/or evolvability can differ markedly between sexes and environments. We discuss the implications of these results for conservation and evolutionary biology. \vspace{5mm} \par\noindent \textbf{Keywords:} Adaptation, Gender load, Monogamy, Mutation load, Sexual conflict.
---
\maketitle\unmarkedfntext{\textsuperscript{§}School of BioSciences, The University of Melbourne, Victoria, Australia.}

@@

Here, we synthesise the empirical literature on sexual selection and population fitness using formal meta-analysis. We focus exclusively on experimental evolution studies that manipulated the presence or strength of sexual selection, and then measured some fitness component, since these provide a particularly strong test of the hypothesis that sexual selection affects the average fitness of populations. Our principle aims were to measure the average net effect of sexual selection on the population-level mean and variance for various fitness components, to test whether this effect varies between stressful and benign environments, and identify key moderators of the effect size of sexual selection treatment. \phantomsection

@@

The grand mean for all effect sizes (**direct, indirect and ambiguous**) was positive (REML $\beta$ = 0.25, 95% CIs: 0.06 to 0.43, $p$ = 0.008; Bayesian $\beta$ = 0.25, 95% CIs: -0.02 to 0.52, $BF_{>0}$ = 32), indicating that sexual selection typically had a net positive effect on the majority of populations and fitness components so far studied. Moreover, the effect sizes associated with the manipulation of sexual selection varied between different fitness traits. Sexual selection had a beneficial effect on most fitness traits, but varied across the three relationships to fitness \autoref{forest_plot}; \href{https://justincally.github.io/SexualSelection/#effect_of_sexual_selection_for_different_fitness_components}{Table S6}). **Sexual selection elevated fitness for traits that shared an ambiguous relationship to fitness (REML $\beta$ = 0.23, 95% CIs: 0.12 to 0.35; Bayesian $\beta$ = 0.23, 95% CIs: 0.03 to 0.41, $BF_{>0}$ = 55, $n$ = 102) and an indirect relationship to fitness (REML $\beta$ = 0.25, 95% CIs: 0.13 to 0.36; Bayesian $\beta$ = 0.24, 95% CIs: 0.04 to 0.42, $BF_{>0}$ = 68, $n$ = 141). Additionally, sexual selection elevated fitness components directly related to fitness, albeit at a lower magnitude (REML $\beta$ = 0.13, 95% CIs: 0.02 to 0.24; Bayesian $\beta$ = 0.13, 95% CIs: -0.07 to 0.3, $BF_{>0}$ = 14, $n$ = 216). An extensive forest plot with predicted effect sizes for each fitness component is presented in Figure S1 and further detailed in model prediction and individual meta-analyses in** \href{https://justincally.github.io/SexualSelection/#effect_of_sexual_selection_for_different_fitness_components}{Table S8 and S9}. Sexual selection significantly reduced two fitness components, namely immunity (REML $\beta$ = -0.42, 95% CIs: -0.64 to -0.20; Bayesian $\beta$ = -0.43, 95% CIs: -0.70 to -0.16, $BF_{>0}$ = 0.0021; $n$ = 35) and body condition (REML $\beta$ = -1.2, 95% CIs: -1.8 to -0.63; Bayesian $\beta$ = -1.2, 95% CIs: -1.8 to -0.61, $BF_{>0}$ = 0.00013, $n$ = 1).  \phantomsection

@@

We found that the sex of the individuals measured (male, female, or a mixture), and the conditions under which the population evolved (stressful or benign) interacted to affect fitness (\hyperlink{page.18}{Table 1, 2}). Sexual selection significantly improved female fitness, and the beneficial effect of sexual selection was significantly stronger for females from populations evolving under stressful conditions (*e.g.* a food source to which they were not well-adapted) than under benign conditions (\autoref{env_sex}a, \hyperlink{page.18}{Table 2}). Sexual selection had a positive but non-significant effect on male fitness, and in contrast to females, fitness benefits were significantly weaker in stressful than benign environments (\autoref{env_sex}a, \hyperlink{page.18}{Table 2}). Consistent with the different consequences of sexual selection for female and male fitness, the mean effect size in mixed-sex samples was non-significantly positive, and there was no significant difference between benign versus stressful conditions (\autoref{env_sex}a, \hyperlink{page.13}{Table 2}). Overall, our results indicate that the positive effect of sexual selection on fitness is greater for females than males, and the difference between the sexes is magnified in stressful environments. **When the lnRR (log Response Ratio) effect size was used to conduct the meta-analysis investigating the impact of sex and environment, the results aligned with those of Hedges' g in that sexual selection elevates population fitness, with its effect magnified for females evolving in stressful environments (Figure S4 and Table S16).**

@@

Our meta-analysis revealed that populations evolving under sexual selection often have higher values for multiple fitness traits, relative to populations where sexual selection was experimentally removed or weakened. **Sexual selection had beneficial effects for traits that were ambiguously or indirectly related to fitness, but had smaller, variable effects on direct measures of fitness such as extinction rate, reproductive success (defined as the number of offspring produced) and the proportion of viable offspring**. Fitness traits related to immunocompetence were an exception: sexual selection typically resulted in weaker immunity. This result is interesting in light of the hypothesised trade-off between sexually-selected phenotypes and immunity, for example, due to immunosuppressive effects of sex hormones [@RN323; @RN324]. Furthermore, the overall benefit of sexual selection was greater for females than males, and this sex difference was magnified in stressful environments. Consistent with stronger selection on female fitness under stress, **female and mixed-sex samples showed substantially reduced phenotypic variance** when sexual selection was applied under stressful as opposed to benign conditions. These results suggest that sexual selection may contribute to population persistence under stressful conditions, such as fluctuating environmental change [@RN173] or spatial variability [@RN189], particularly since female reproductive output is often a limiting factor in population growth [@RN188].

The results of the meta-analysis support predictions that sexual selection can improve population fitness and accelerate adaptation [@RN94; @RN173; @RN54; @RN141; @RN230; @RN198; @RN58]. Specifically, our results support arguments that positive genetic correlations between male mating success and female non-sexual traits are common, allowing females to benefit from a genome that has been 'purged' of deleterious alleles through selection on males [@RN94; @RN198; @RN95]. What is less clear is why sexual selection had a larger effect on female trait means and variances as opposed to males (the sex in which sexual selection is typically strongest). Below we discuss possible explanations for this result, in light of the core principle that the speed and magnitude of adaptation depends on additive genetic variance, the structure of genetic covariance, and the strength of selection [@RN317; @RN318].

@@

From each paper, we first attempted to extract the arithmetic means, standard deviations and sample sizes of each of the different treatment groups, which facilitate calculation of effect size (see below). Typically, there were two or three treatments, which varied in the strength of sexual selection. For some papers, summary statistics were not written down, but were presented in a figure such as a bar chart: in these cases, we extracted the data using *WebPlotDigitizer* v.3.12 [@RN195]. If the treatment means were not reported (and the raw data were unavailable), we instead calculated effect size from test statistics comparing treatment means (*e.g.*, $F$, $t$, $z$ or $\chi^2$ values), which we used to estimate effect size using several formulae (see below).

@@

For the meta-analysis testing whether sexual selection affects phenotypic variance (as opposed to the mean), we estimated the difference in variance between each pair of treatments using the natural logarithm of the ratio between the coefficient of variation for each group (termed *lnCVR*) [@RN177].**The use of _lnCVR_ allows us to determine the effects of sexual selection on phenotypic variance, with the coefficient of variation implicitly controlling for the mean-variance relationship seen in the dataset (FIG SXX). As a supplement, we also calculated the natural logarithm of the absolute ratio between the absolute variation for each group (_lnVR_) in order to assess the impact of sexual selection on trait variance, irrespective of their magnitudes [@RN177].** The calculation of *lnCVR* relies on the availability of arithmetic means, standard deviations and sample sizes for the two treatment groups [@RN177; @RN190], and so we were only able to calculate *lnCVR* for 352 of 459 comparisons.  \phantomsection

@@

![Tests of publication bias are mixed and suggest publication bias may be present. (**a**) Inspection and statistical tests of the funnel plot reveal large amounts of heterogeneity in the dataset with asymmetry from increased low-powered, large effect studies. (**b**) No significant correlation exists between journal impact factor and effect size, (**c**) additionally there is no significant correlation between effect size and year of publication when testing for the time-lag bias. For (**b**) and (**c**) point size is proportional to the precision of the effect size (*i.e.* the inverse of its variance). Grey envelopes in (**b**) and (**c**) represent the 95 % CI of the linear regression.\label{publication_bias}](Figures/publicationbias.pdf)

>>>>

---
title: '\huge{Sexual selection improves population fitness: a systematic review and meta-analysis}'
author: 
- Justin G. Cally\textsuperscript{§*}, Devi Stuart-Fox\textsuperscript{§} and Luke Holman\textsuperscript{§}
- 
- '\textsuperscript{*}\textit{justin.g.cally@gmail.com}'
output:
  pdf_document: default
  html_document: default
  word_document:
      pandoc_args: ["--smart"]
bibliography: Bibliography/Bibliography.bib
subparagraph: yes
csl: Bibliography/NComms.csl
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[LO,LE]{\textsl{\leftmark}}
  \rhead[]{Sexual selection improves fitness} \usepackage{titlefoot}
  \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno} \linenumbers \usepackage{setspace}\onehalfspacing
    \usepackage{titlesec} \titleformat{\section}{\normalfont\scshape\LARGE}{\thesection}{1em}{} \titleformat{\subsection}{\normalfont\Large}{\thesubsection}{1em}{} \titleformat{\subsubsection}{\normalfont\large}{\thesubsubsection}{1em}{} \usepackage{abstract} \renewcommand{\abstractnamefont}{\normalfont\Large} 
    \usepackage{hyperref}\hypersetup{colorlinks=false, linkcolor=blue, filecolor=blue, urlcolor=blue} 
    \usepackage{color,soul}
    \raggedbottom    
abstract: Sexual selection has manifold ecological and evolutionary consequences, making its net effect on population fitness difficult to predict from first principles. Instead, one can empirically determine whether sexual selection increases or decreases population fitness by  manipulating the opportunity for sexual selection in replicate populations, allow evolution to proceed, and then compare the evolved average fitness between the treatment groups. Here, we synthesise 459 effect sizes from 65 pertinent experimental evolution studies using meta-analysis. We find that sexual selection **on males** tends to elevate the mean and reduce the variance for many fitness traits, especially in females, and for populations evolving under stressful conditions. We conclude that the beneficial population-level consequences of sexual selection typically outweigh the harmful ones and that the effects of sexual selection on total selection, genetic variance, and/or evolvability can differ markedly between sexes and environments. We discuss the implications of these results for conservation and evolutionary biology. \vspace{5mm} \par\noindent \textbf{Keywords:} Adaptation, Gender load, Monogamy, Mutation load, Sexual conflict.
---
\maketitle\unmarkedfntext{\textsuperscript{§}School of BioSciences, The University of Melbourne, Victoria, Australia.}

@@

Here, we synthesise the empirical literature on sexual selection and population fitness using formal meta-analysis. We focus exclusively on experimental evolution studies that manipulated the presence or strength of sexual selection **on males**, and then measured some fitness component, since these provide a particularly strong test of the hypothesis that sexual selection affects the average fitness of populations. Our principle aims were to measure the average net effect of sexual selection on the population-level mean and variance for various fitness components, to test whether this effect varies between stressful and benign environments, and identify key moderators of the effect size of sexual selection treatment. \phantomsection

@@

The grand mean for all effect sizes (**direct, indirect and ambiguous**) was positive (REML $\beta$ = 0.25, 95% CIs: 0.06 to 0.43, $p$ = 0.008; Bayesian $\beta$ = 0.25, 95% CIs: -0.02 to 0.52, $BF_{>0}$ = 32), indicating that sexual selection **on males** typically had a net positive effect on the majority of populations and fitness components so far studied. Moreover, the effect sizes associated with the manipulation of sexual selection varied between different fitness traits. Sexual selection had a beneficial effect on most fitness traits, but varied across the three relationships to fitness \autoref{forest_plot}; \href{https://justincally.github.io/SexualSelection/#effect_of_sexual_selection_for_different_fitness_components}{Table S6}). **Sexual selection elevated fitness for traits that shared an ambiguous relationship to fitness (REML $\beta$ = 0.23, 95% CIs: 0.12 to 0.35; Bayesian $\beta$ = 0.23, 95% CIs: 0.03 to 0.41, $BF_{>0}$ = 55, $n$ = 102) and an indirect relationship to fitness (REML $\beta$ = 0.25, 95% CIs: 0.13 to 0.36; Bayesian $\beta$ = 0.24, 95% CIs: 0.04 to 0.42, $BF_{>0}$ = 68, $n$ = 141). Additionally, sexual selection elevated fitness components directly related to fitness, albeit at a lower magnitude (REML $\beta$ = 0.13, 95% CIs: 0.02 to 0.24; Bayesian $\beta$ = 0.13, 95% CIs: -0.07 to 0.3, $BF_{>0}$ = 14, $n$ = 216). An extensive forest plot with predicted effect sizes for each fitness component is presented in Figure S1 and further detailed in model prediction and individual meta-analyses in** \href{https://justincally.github.io/SexualSelection/#effect_of_sexual_selection_for_different_fitness_components}{Table S8 and S9}. Sexual selection significantly reduced two fitness components, namely immunity (REML $\beta$ = -0.42, 95% CIs: -0.64 to -0.20; Bayesian $\beta$ = -0.43, 95% CIs: -0.70 to -0.16, $BF_{>0}$ = 0.0021; $n$ = 35) and body condition (REML $\beta$ = -1.2, 95% CIs: -1.8 to -0.63; Bayesian $\beta$ = -1.2, 95% CIs: -1.8 to -0.61, $BF_{>0}$ = 0.00013, $n$ = 1).  \phantomsection

@@

We found that the sex of the individuals measured (male, female, or a mixture), and the conditions under which the population evolved (stressful or benign) interacted to affect fitness (\hyperlink{page.18}{Table 1, 2}). Sexual selection **on males** significantly improved female fitness, and the beneficial effect of sexual selection was significantly stronger for females from populations evolving under stressful conditions (*e.g.* a food source to which they were not well-adapted) than under benign conditions (\autoref{env_sex}a, \hyperlink{page.18}{Table 2}). Sexual selection had a positive but non-significant effect on male fitness, and in contrast to females, fitness benefits were significantly weaker in stressful than benign environments (\autoref{env_sex}a, \hyperlink{page.18}{Table 2}). Consistent with the different consequences of sexual selection for female and male fitness, the mean effect size in mixed-sex samples was non-significantly positive, and there was no significant difference between benign versus stressful conditions (\autoref{env_sex}a, \hyperlink{page.13}{Table 2}). Overall, our results indicate that the positive effect of sexual selection on fitness is greater for females than males, and the difference between the sexes is magnified in stressful environments. **When the lnRR (log Response Ratio) effect size was used to conduct the meta-analysis investigating the impact of sex and environment, the results aligned with those of Hedges' g in that sexual selection elevates population fitness, with its effect magnified for females evolving in stressful environments (Figure S4 and Table S16).**

@@

Our meta-analysis revealed that populations evolving under sexual selection often have higher values for multiple fitness traits, relative to populations where sexual selection **on males** was experimentally removed or weakened. **Sexual selection had beneficial effects for traits that were ambiguously or indirectly related to fitness, but had smaller, variable effects on direct measures of fitness such as extinction rate, reproductive success (defined as the number of offspring produced) and the proportion of viable offspring**. Fitness traits related to immunocompetence were an exception: sexual selection typically resulted in weaker immunity. This result is interesting in light of the hypothesised trade-off between sexually-selected phenotypes and immunity, for example, due to immunosuppressive effects of sex hormones [@RN323; @RN324]. Furthermore, the overall benefit of sexual selection was greater for females than males, and this sex difference was magnified in stressful environments. Consistent with stronger selection on female fitness under stress, **female and mixed-sex samples showed substantially reduced phenotypic variance** when sexual selection was applied under stressful as opposed to benign conditions. These results suggest that sexual selection may contribute to population persistence under stressful conditions, such as fluctuating environmental change [@RN173] or spatial variability [@RN189], particularly since female reproductive output is often a limiting factor in population growth [@RN188].

The results of the meta-analysis support predictions that sexual selection **on males** can improve population fitness and accelerate adaptation [@RN94; @RN173; @RN54; @RN141; @RN230; @RN198; @RN58]. Specifically, our results support arguments that positive genetic correlations between male mating success and female non-sexual traits are common, allowing females to benefit from a genome that has been 'purged' of deleterious alleles through selection on males [@RN94; @RN198; @RN95]. What is less clear is why sexual selection had a larger effect on female trait means and variances as opposed to males (the sex in which sexual selection is typically strongest). Below we discuss possible explanations for this result, in light of the core principle that the speed and magnitude of adaptation depends on additive genetic variance, the structure of genetic covariance, and the strength of selection [@RN317; @RN318].

@@

From each paper, we first attempted to extract the arithmetic means, standard deviations and sample sizes of each of the different treatment groups, which facilitate calculation of effect size (see below). Typically, there were two or three treatments, which varied in the strength of sexual selection **on males through manipulations to the adult sex ratio, in these cases we considered treatments with the greater male to female ratio to be the high sexual selection treatment group**. For some papers, summary statistics were not written down, but were presented in a figure such as a bar chart: in these cases, we extracted the data using *WebPlotDigitizer* v.3.12 [@RN195]. If the treatment means were not reported (and the raw data were unavailable), we instead calculated effect size from test statistics comparing treatment means (*e.g.*, $F$, $t$, $z$ or $\chi^2$ values), which we used to estimate effect size using several formulae (see below).

@@

For the meta-analysis testing whether sexual selection affects phenotypic variance (as opposed to the mean), we estimated the difference in variance between each pair of treatments using the natural logarithm of the ratio between the coefficient of variation for each group (termed *lnCVR*) [@RN177]: $ln(CVfitness_{SS\ high}/CVfitness_{SS\ low})$ .**The use of _lnCVR_ allows us to determine the effects of sexual selection on phenotypic variance, with the coefficient of variation implicitly controlling for the mean-variance relationship seen in the dataset (Figure S6). As a supplement, we also calculated the natural logarithm of the absolute ratio between the absolute variation for each group (_lnVR_) in order to assess the impact of sexual selection on trait variance, irrespective of their magnitudes [@RN177].** The calculation of *lnCVR* relies on the availability of arithmetic means, standard deviations and sample sizes for the two treatment groups [@RN177; @RN190], and so we were only able to calculate *lnCVR* for 352 of 459 comparisons.  \phantomsection

@@

![Tests of publication bias are mixed and suggest publication bias may be present. (**a**) Inspection and statistical tests of the funnel plot reveal large amounts of heterogeneity in the dataset with asymmetry from increased low-powered, large effect studies. (**b**) No significant correlation exists between journal impact factor and effect size, (**c**) additionally there is no significant correlation between effect size and year of publication when testing for the time-lag bias. For (**b**) and (**c**) point size is proportional to the precision of the effect size (*i.e.* the inverse of its variance). For (**a**), the dashed red lines represent the 95 % CIs of the grand mean estimate for all effect sizes and the black dotted and dashed lines depict the 95 % and 99 % CIs for the dataset. The Grey envelopes in (**b**) and (**c**) represent the 95 % CI of the linear regression.\label{publication_bias}](Figures/publicationbias.pdf)
